---
title: "Peptide workflow with DaparToolshed"
author: 
- name: Manon Gaudin
- name: Nicolas Enjalbert Courrech 
- name: Samuel Wieczorek
- name: Pierre Neuvial
- name: Thomas Burger
package: DaparToolshed
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        self_contained: false
        highlight: tango
        toc_float: true
bibliography: references.bib
vignette: >
    %\VignetteIndexEntry{Peptide workflow with DaparToolshed}
    %\VignetteEngine{knitr::rmarkdown}
    %%\VignetteKeywords{Mass Spectrometry, Quantitative}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setups, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```


# Introduction

`DaparToolshed` is a package for analyzing quantitative data from label-free bottom-up proteomics experiments. It uses `MultiAssayExperiment`, `SummarizedExperiment` and `QFeatures` data structures (@gatto_qfeatures_2025).

Many R packages or suites of packages are available to process mass spectrometry-based proteomics data: [R for mass spectrometry](https://www.rformassspectrometry.org/), [MSstat](https://msstats.org/), [DeqMS](https://www.bioconductor.org/packages/devel/bioc/html/DEqMS.html), [MSqRob](https://bioconductor.org/packages/release/bioc/html/msqrob2.html), [msmsEDA](https://www.bioconductor.org/packages/release/bioc/html/msmsEDA.html)/[msmsTests](https://www.bioconductor.org/packages/release/bioc/html/msmsTests.html), [proDA](https://www.bioconductor.org/packages/release/bioc/html/proDA.html), [scp](https://www.bioconductor.org/packages/release/bioc/html/scp.html), [DEP](https://www.bioconductor.org/packages/release/bioc/html/DEP.html), and many others. They tend to share (i) statistical methodologies / algorithms (about normalisation, imputation, differential analysis and so forth) and (ii) Bioconductor formats (e.g., `QFeatures`) or low levels routines (e.g., [MsCoreUtils](https://www.bioconductor.org/packages/release/bioc/html/MsCoreUtils.html)); however, the redundancy makes it possible to find a diversity of workflows that match the specificities of each experimental design. In this landscape, `DaparToolshed` is an update of the decade-old package `DAPAR`, from the [DAPAR](https://www.bioconductor.org/packages/release/bioc/html/DAPAR.html)/[Prostar](https://www.bioconductor.org/packages/release/bioc/html/Prostar.html) suite (@Wieczorek2017-nn), which makes it possible to leverage the extended capabilities of the `QFeatures` format. However, as `DAPAR`  leverages data structure from [MSnbase](https://www.bioconductor.org/packages/release/bioc/html/MSnbase.html) that are still used nowdays in spite of `QFeatures` advent, we propose separated packages for sakes of simpler use and simpler maintenance.

Starting from a table of abundance values and associated metdata, `DaparToolshed` proposes functions to build a complete statistical analysis workflow up to the selection of differentially expressed proteins in the contrasted conditions. `DaparToolshed` makes it possible to import data at different levels: precursor, peptide or protein, each requiring specific processing.

This vignette describes a complete precursor- or peptide-level workflow.

This package is used by the `Prostar2` package, which provides a user-friendly graphical interfaces, so that programming skill is not required to achieve the data analysis.

## Installation

To install `DaparToolshed` :

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("DaparToolshed", quietly = TRUE)) pak::pak("edyp-lab/DaparToolshed")
if (!requireNamespace("Pirat", quietly = TRUE)) pak::pak("Pirat")
```

```{r setup, message = FALSE}
library(DaparToolshed)
```


# Import dataset

For testing purposes, the `DaparToolshedData` package is accompanied with several example datasets.

The dataset used in this vignette is an extract from the `dou2019_mouse` dataset in the `scpdata` package.

```{r}
# Library for example dataset
if (!requireNamespace("scpdata", quietly = TRUE)) pak::pak("scpdata")

library(scpdata)
```
We illustrate this pipeline on single-cell data to get statistical power but the pipeline is also available on bulk proteomic data (considering cells and samples are equivalent in this analysis). 

The dataset is composed of 15,492 peptides and 72 cells, equally divided into three conditions ("c10", "SVEC" and "RAW").
To obtain more information on this dataset please refer to the help of the function `?scpdata::dou2019_mouse` or the article @dou2019_mousedata.


```{r dataset}
dta <- dou2019_mouse()
## PREPARE DATA -----
# keep only wanted conditions
dtacond <- c("C10", "SVEC", "RAW")
colselect <- which(SummarizedExperiment::colData(dta)$SampleType %in% dtacond)

# create info for coldata
nomcoldta <- colnames(SummarizedExperiment::assay(dta[[13]][, colselect]))
nomcond <- SummarizedExperiment::colData(dta)[colselect, "SampleType"]
# met tt ensemble
sampledta <- data.frame(
  Sample.name = nomcoldta,
  Condition = nomcond,
  Bio.Rep = seq_along(nomcoldta)
)
sampledta$Sample.name <- gsub(".", "_", sampledta$Sample.name, fixed = TRUE)
sampledta <- sampledta[order(sampledta$Condition), ]


# select assay
assaydta <- SummarizedExperiment::assay(dta[[13]][, colselect])
colnames(assaydta) <- gsub(".", "_", colnames(assaydta), fixed = TRUE)
rownames(assaydta) <- seq_len(nrow(assaydta))

# get rowdata
rowdta <- SummarizedExperiment::rowData(dta[[13]])
rownames(rowdta) <- seq_len(nrow(rowdta))

# create columns for metacell
metadta <- as.data.frame(replicate(ncol(assaydta),
  rep("By MS/MS", nrow(assaydta)),
  simplify = FALSE
))
metadta <- setNames(metadta, paste0("Id_", colnames(assaydta)))

# merge together
quantdta <- cbind(assaydta, rowdta, metadta)
quantdta <- as.data.frame(quantdta)
colnames(quantdta) <- gsub(".", "_", colnames(quantdta), fixed = TRUE)
```

`DaparToolshed` uses `MultiAssayExperiment`, `SummarizedExperiment` and `QFeatures` data structures.

The function `createQFeatures()` allows for the importation of a given dataset to a `QFeatures` format. The dataset can come from any tabulated-like file containing quantitative data and metadata, accompanied by another dataframe for the sample description. The `QFeatures` format allows you to keep track of the various transformations made at each step.

For each peptide (or precursor) in each sample of the dataset, it is possible to tag the evidence that links the abundance to the molecule identified: for instance, if the abundance value is missing or imputed, if the quantification-identification match result from a transfer between samples or if there has been a fragmentation spectrum identified in relation to the quantified elution profile, etc. If such metadata is available, then the corresponding column indexes can be indicated in `indexForMetacell`. This allows the information to be leveraged during the workflow. 

The quantitative data do not need to be logged beforehand, as setting `logData = TRUE` will yield automatic log-transform. Similarly, with `force.na = TRUE`, any quantitative data with a null value or a NaN will be converted to NA. It is necessary to specify the column of the table where the mother proteins are indicated (`'parentProtId'` parameter) as an adjacency matrix is created when the function is executed. This matrix provides easy access to peptide-protein relationships, including shared peptides, i.e., when the same peptide belongs to several proteins. To run a peptidomics analysis (i.e., peptides are analysed irrespective of their mother proteins), as no peptide-to-protein aggregation is necessary, it is more suited to rely on the same workflow as for protein-level analysis.

To decrease the run time of the vignette, we only use the first 1000 peptides. The analyse should be on all the peptides. 

```{r importdata, message = FALSE}
obj <- createQFeatures(
  data = quantdta,
  file = "dou2019_mouse",
  sample = sampledta,
  indQData = 1:72,
  keyId = "sequence",
  indexForMetacell = 87:158,
  logData = TRUE,
  force.na = TRUE,
  typeDataset = "peptide",
  parentProtId = "DatabaseAccess",
  analysis = "Pept_Data",
  software = "maxquant"
)

obj <- obj[1:1000, ]

obj
```

```{r}
design_qf(obj)
```

The data obtained includes two assays: `original` and `logAssay`, respectively for the non-logged data and the logged data. This shows how the `QFeatures` format allows for recording the state of the dataset after each step. 


# Filtering

The first step in this workflow is peptide filtering.  Not all peptides in the dataset are relevant for analysis. Some may be contaminants, for example, while others may have too many missing values to be taken into account.
This filtering can be done using information contained in quantitative data tags (see above) or metadata, i.e., all the variables from the original dataset that were not quantitative data and can be accessed through the `SummarizedExperiment::rowData()` function. 

```{r metadataaccess, message = FALSE}
# quantitative data tags
head(qMetacell(obj[[length(obj)]]))

# metadata
head(SummarizedExperiment::rowData(obj[[length(obj)]]), n = 3)
```

To apply a filter based on tags, the `FunctionFilter()` function should be used. The functions `QFeatures::filterFeature()` and `QFeatures::VariableFilter()` can be used to create filters from metadata.

```{r filters, message = FALSE}
# create filter to remove empty lines
filter_emptyline <- FunctionFilter("qMetacellWholeLine",
  cmd = "delete",
  pattern = "Missing MEC"
)
# filter lines with too many NA is every condition
filter_manyNA <- FunctionFilter("qMetacellOnConditions",
  cmd = "delete",
  mode = "AllCond",
  pattern = c("Missing MEC", "Missing POV"),
  conds = SummarizedExperiment::colData(obj)$Condition,
  percent = TRUE,
  th = 0.8,
  operator = ">"
)
# create filter to remove contaminant
filter_contaminant <- QFeatures::VariableFilter(
  field = "isContaminant",
  value = "TRUE",
  condition = "==",
  not = TRUE
)
```

Doing so creates the filter but does not apply them to the data. All created filters can be applied at once using the `filterFeaturesOneSE()` function, which will perform filtering on the assay indicated by `i` and create a new assay with the filtered peptides.

```{r filtering, message = FALSE}
# apply all filters and create new assay
obj <- filterFeaturesOneSE(
  object = obj,
  i = length(obj),
  name = "Filtered",
  filters = list(filter_emptyline, filter_manyNA, filter_contaminant)
)
```

It is advised to remove proteins that no longer have an associated peptide following this filtering from the adjacency matrix in order to facilitate the subsequent aggregation step.

```{r removeprot, message = FALSE}
# remove proteins with no associated peptides
X <- QFeatures::adjacencyMatrix(obj[[length(obj)]])
SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix <-
  X[, -which(Matrix::colSums(X) == 0)]

obj
```


# Normalization

The normalization step reduces biases introduced at any preliminary stage, as to make samples more comparable. Though not a full batch effect correction method (which can be required depending on the experimental design), it can limit some simple batch effects. During this step, peptide abundances are essentially rescaled between samples of the same conditions or of all conditions.

Several algorithms are implemented in `DaparToolshed` and the list of available methods can be accessed by using `normalizeMethods()` :

```{r normalizationmethods}
# list of available methods
normalizeMethods()
```

The methods proposed acts as described below:

- `GlobalQuantileAlignment` : Aligns the quantiles of intensity distributions across all samples. (This method should be used cautiously, as it imposes a high-impact normalization. Broadly speaking, the abundances are replaced by their order statistics).
- `SumByColumns` : Normalizes each abundance value by the total abundance of its corresponding sample. Un-logged intensities are used for processing. This method assumes the total amount of biological material is equal in all the samples.
- `QuantileCentering` : Aligns the intensity distributions to a specific quantile, e.g., the median or a lower quantile, either across all samples or within each condition.
- `MeanCentering` : Aligns the intensity distributions to their means, either across all samples or within each condition. Optionally, unit variance can be enforced.
- `LOESS` : Applies a cyclic LOESS normalization, either across all samples or within each condition. The intensity values are normalized by means of a local regression model of the difference of intensities as function of the median intensity value (@cleveland_robust_1979) (see @gentleman_limma:_2005 for implementation details).
- `vsn` : Applies the Variance Stabilization Normalization method, either across all samples or within each condition (@huber_variance_2002).


The `normalizeFunction()` function allows you to normalize data using any of the methods described above and create a new assay with normalized data. The parameters to be defined depend on the chosen method. The `type` argument allows to indicate whether the method is applied to the entire dataset at once (`"overall"`) or whether each condition is normalized independently (`"within conditions"`).

```{r normalization}
obj <- normalizeFunction(obj,
  method = "MeanCentering",
  scaling = TRUE,
  type = "overall"
)

obj
```

The `compareNormalizationD_HC()` function provides a way to visually compare the quantitative data before and after normalization. This plot shows the influence of the normalization method.

```{r normalizationplot, message = FALSE, warning=FALSE}
obj1 <- obj[[length(obj)]]
obj2 <- obj[[length(obj) - 1]]
protId <- DaparToolshed::idcol(obj1)

.n <- floor(0.02 * nrow(obj1))
.subset <- seq_len(nrow(obj1))

compareNormalizationD_HC(
  qDataBefore = SummarizedExperiment::assay(obj1),
  qDataAfter = SummarizedExperiment::assay(obj2),
  keyId = SummarizedExperiment::rowData(obj1)[, protId],
  conds = DaparToolshed::design_qf(obj)$Condition,
  n = .n,
  subset.view = .subset
)
```


# Imputation

It is common to face a significant number of missing values. To overcome this problem, it is often necessary to resort to imputation. However, it is important to note that data imputation amount to creating experimental measurements out of thin air (and of a bit of mathematics too). It is thus recommended to avoid imputation whenever possible. However, most algorithms required subsequently cannot deal with missing values, and the few that claim to be robust to missing values either relies on implicit or explicit imputation, which makes them no better.

First, it is interesting to look at the quantity and distribution of missing values. The `metacellPerLinesHisto_HC()` and `metacellPerLinesHistoPerCondition_HC()` functions give access to the number of rows containing a given number of missing values, respectively for all samples and within each condition. These functions use the quantitative data tags described above.

```{r missingvalueplots, warning=FALSE}
pal <- unique(GetColorsForConditions(design_qf(obj)$Condition))
pattern <- c("Missing MEC", "Missing POV")
grp <- design_qf(obj)$Condition

# number of line with different amount of NA
metacellPerLinesHisto_HC(obj[[length(obj)]],
  group = grp,
  pattern = pattern
)
# number of line with different amount of NA per condition
metacellPerLinesHistoPerCondition_HC(obj[[length(obj)]],
  group = grp,
  pattern = pattern
)
```

The `hc_mvTypePlot2()` function displays density plots showing the distribution of partially observed values for each condition. The x-axis corresponds to the mean intensity of a peptide in a condition, while the y-axis indicates the number of observed values for that peptide under the same condition.

It can be observed that the distribution tends to shift towards higher values when fewer missing values are present. This is a sign that at least some of the missing values originate from peptides that are below the lower limit of the mass spectrometer.

```{r missingvaluedistribplots}
hc_mvTypePlot2(obj[[length(obj)]], group = grp, pattern = pattern, pal = pal)
```

Numerous imputation methods exist. Here, data imputation is performed using the `wrapperPirat()` function, which is a wrapper for the `Pirat` method. `Pirat` is an imputation method that uses a penalized maximum likelihood strategy and accounts for sibling peptide correlations (@etourneau_penalized_2024). It does not require any parameter tuning. For more information on this method, see the [Pirat](https://www.bioconductor.org/packages/release/bioc/html/Pirat.html) package.

```{r imputation, message = FALSE}
obj <- wrapperPirat(
  data = obj,
  adjmat = SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix,
  extension = "base"
)

obj
```


# Aggregation

The next step is to aggregate the peptide-level data into protein-level data. For optimal results in the subsequent differential analysis, it is important to aggregate only after imputation. 

Beforehand, it may be useful to look at the dataset in more details and to choose the most appropriate aggregation method accordingly. The `getProteinsStats()` function displays various information about peptides and associated proteins from the adjacency matrix.

```{r aggregationadjmatrix}
X <- SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix
info_pept <- getProteinsStats(X)

message(
  "Total number of peptides : ", info_pept$nbPeptides, "\n",
  "Number of specific peptides : ", info_pept$nbSpecificPeptides, "\n",
  "Number of shared peptides : ", info_pept$nbSharedPeptides, "\n"
)

message(
  "Total number of proteins : ", info_pept$nbProt, "\n",
  "Number of protein with only specific peptides : ",
  length(info_pept$protOnlyUniquePep), "\n",
  "Number of protein with only shared peptides : ",
  length(info_pept$protOnlySharedPep), "\n",
  "Number of protein with both : ", length(info_pept$protMixPep), "\n"
)
```

There are predominantly specific peptides, but several proteins contain shared peptides. The question of how to manage these shared peptides may arise. Contrarily to most aggregation functions, `RunAggregation()` allows shared peptides to be taken into account in various ways --to be specified using the `includeSharedPeptides` argument. First, they can be handled as specific peptides (`Yes_As_Specific`), in which case they will be duplicated for each protein they belong to. Second, they can have their intensity redistributed proportionally across the different proteins (`Yes_Iterative_Redistribution` or `Yes_Simple_Redistribution`). Finally, shared peptides can be ignored (`No`). If `Yes_Iterative_Redistribution`, `Yes_Simple_Redistribution` or `No` is chosen, proteins containing only shared peptides will have no associated value. 
In addition, there is the choice of aggregation function, used for quantitative feature aggregation, with `operator`. This can be either `Sum`, `Mean`, `Median`, `medianPolish` or `robustSummary`.

```{r aggregation, message = FALSE}
obj <- RunAggregation(obj,
  adjMatrix = "adjacencyMatrix",
  includeSharedPeptides = "Yes_Iterative_Redistribution",
  operator = "Mean",
  considerPeptides = "allPeptides",
  ponderation = "Global",
  max_iter = 500
)

obj
```

After aggregation, it can be seen that there are still some missing values. These come from proteins that only have shared peptides, which have been left out using this method, as explained above. 

```{r aggregationsummary}
summary(SummarizedExperiment::assay(obj[["aggregated"]]))
```

A new filter can thus be applied to these protein-level data to remove them.

```{r filteringprot}
# apply the filter to remove empty lines created
obj <- filterFeaturesOneSE(
  object = obj,
  i = length(obj),
  name = "FilterProt",
  filters = list(filter_emptyline)
)

obj
```


# Differential Analysis

Differential protein expression analysis compares the relative abundance of the same protein between two conditions. The comparison between two conditions results in what is called a contrast, for example “A vs. B.” 
The dataset contains three conditions. To illustrate the differential analysis, we only compare two conditions : "C10" vs "RAW".

```{r}
# selectionne 2 conditions
indexcond <- which(colData(obj)$Condition %in% c("C10", "RAW"))
obj_selectcond <- obj[, indexcond]
```



A common practice is to perform a test between two conditions for each protein and plot a volcano plot (a graph with log fold change on the x-axis and p-values on the y-axis on a $-\log_{10}$ scale). Conventional differential analysis seeks to refine a selection of proteins on a volcano plot by setting thresholds based on (log) $p$-values and (log) fold changes. This selection procedure does not provide guarantees regarding the FDR (False Discovery Rate) @ebrahimpoor2021inflated. However, \emph{post hoc} methods @GS2011 allow for control of the False Discovery Proportion (FDP, mathematically $FDR = \mathbb{E}[FDP]$). These methods have the property of upper bounding the FDP for *arbitrary, multiple, simultaneous, and possibly data-driven subsets* of proteins with **statistical confidence**. In particular, these FDP bounds are valid for protein selections derived from the visual (and possibly iterative) analysis of a volcano plot.

For this analysis we use the \texttt{sanssouci} package @sansSouci-package implementing \emph{post hoc} methods developed in @blanchard20post-hoc. 
The default method is called the Simes method. When the number of samples per condition is large enough, a more powerful method can be obtained by a calibration principle introduced in @blanchard20post-hoc: it relies on permutations of condition labels to learn the dependence between proteins. By construction, this method is more computationally intensive than the baseline method. 

### Initialisation of SansSouci object

The first step is to init a `SansSouci` object. More precisely, we init a `SansSouci4DT` object which provide a `SansSouci` object for the `DaparToolshed` pipeline described in this vignette. 

```{r}
if (!require(sanssouci4DaparToolshed)) {
  pak::pak("sanssouci-org/sanssouci4DaparToolshed")
}
library(sanssouci4DaparToolshed)

if (!require(sanssouci)) {
  pak::pak("sanssouci-org/sanssouci")
}
library(sanssouci)

sanssouci_object <- SansSouci4DT(obj_selectcond)
```

### Hypothesis testing and post hoc calibration 

To compute tests, we need to choose some parameters. 
The parameter `alpha` is the error level such that the confidence level $1-\alpha$ corresponds to the probability of obtaining an uper bound on the FDP. To approximate the FDR, we propose to choose `alpha = 0.5$ corresponding to the median of the FDP. 

The statistical test and the number of permutations $B$ depend on the number $n$ of samples per condition @enjalbert2022powerful:

- If $n \geq 10$, then use `rowTestFUN = sanssouci::rowWelchTests` and `B = 1000`. The calibrated Simes method will be used. 
- If $n \leq 10$, then replace the Welch $t$-test by a moderated test, like `limma`~@limma, using the parameters `rowTestFUN = sanssouci4DaparToolshed::rowLimmaTest` and `B = 0`. The Simes method will be used.

The running example data thus correspond to the first case. 


```{r}
alpha <- 0.5
rowTestFUN <- sanssouci::rowWelchTests
B <- 1000
```

To compute p-values, fold change and the post hoc bounds calibration, please use `sanssouci::fit()` function.

```{r}
sanssouci_object_cal <- fit(sanssouci_object,
  alpha = alpha,
  rowTestFUN = rowTestFUN, B = B
)
sanssouci_object_cal
```

From the `sanssouci_object_cal`, the p-values and the log fold change can be extract using, respectively, `sanssouci::pValues()` and `sanssouci::foldChanges()` functions.

```{r}
pval <- as.vector(pValues(sanssouci_object_cal))
logFC <- as.vector(foldChanges(sanssouci_object_cal))
```


### Volcano plot and post hoc bound

Then, the volcano plot can be drawn. You can make any selection you want, *post hoc* method provides statistical guaranties.   

```{r}
pval_threshold <- 0.01
logfc_threshold <- 0.04

volcanoPlot(
  sanssouci_object_cal,
  pval_thr = pval_threshold,
  logfc_thr = logfc_threshold, interactive = FALSE
)
```

In our case, thresholds allow for the selection of 40 proteins. Among these proteins, the 50% statistical confidence level indicates that there are at least 35 proteins that are actually differentially expressed. This means that the proportion of false positives does not exceed 12%.

You can iteratively refine the thresholds as many times as necessary. The statistical guarantees will remain intact. 

### Extraction of selected proteins

You can save the selection in the `QFeatures` object with the `export2dapartoolshed` function. Please use the fitted `SansSouci4DT` object and provide the thresholds on p-values and log fold change. 

```{r}
obj_selectcond <- export2dapartoolshed(
  sanssouci_object_cal,
  pval_thr = pval_threshold,
  logfc_thr = logfc_threshold,
  qfeature_obj = obj_selectcond
)
```

This function store a `data.frame` in the metadata of the `QFeatures` object. 

```{r}
head(metadata(obj_selectcond)$posthoc_selection)
```

This `data.frame contains at least three columns: the protein names, the $p$-values and the log fold changes. 
In addition, each of the following columns contains a user-defined selection (the column name begins by "sel\_"). This binary vector indicates whether the protein was selected. The statistics of each selection can be accessed using the `getPostHocBound()` function, as follows: 

```{r}
getPostHocBound(obj_selectcond, "sel_1")
```


## For more complex experimental design

If your experimental design is more complex than two conditions (for exemple, more than two conditions or with covariates), please use post hoc method for linear model developped by @davenport2025fdp. This method is implemented in the package `sanssouci`. For now, please follow the [vignette for linear model](https://sanssouci-org.github.io/sanssouci/articles/post-hoc_differential-expression_linear-model.html).

# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
