---
title: "Peptide workflow with DaparToolshed"
author: 
- name: Manon Gaudin
- name: Nicolas Enjalbert Courrech 
- name: Samuel Wieczorek
- name: Pierre Neuvial
- name: Thomas Burger
package: DaparToolshed
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        self_contained: false
        highlight: tango
        toc_float: true
bibliography: references.bib
vignette: >
    %\VignetteIndexEntry{Peptide workflow with DaparToolshed}
    %\VignetteEngine{knitr::rmarkdown}
    %%\VignetteKeywords{Mass Spectrometry, Quantitative}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setups, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```


# Introduction

`DaparToolshed` is a package for analyzing quantitative data from label-free bottom-up proteomics experiments. It uses `MultiAssayExperiment`, `SummarizedExperiment` and `QFeatures` data structures (@gatto_qfeatures_2025).

Many R packages or suites of packages are available to process mass spectrometry-based proteomics data: [R for mass spectrometry](https://www.rformassspectrometry.org/), [MSstat](https://msstats.org/), [DeqMS](https://www.bioconductor.org/packages/devel/bioc/html/DEqMS.html), [MSqRob](https://bioconductor.org/packages/release/bioc/html/msqrob2.html), [msmsEDA](https://www.bioconductor.org/packages/release/bioc/html/msmsEDA.html)/[msmsTests](https://www.bioconductor.org/packages/release/bioc/html/msmsTests.html), [proDA](https://www.bioconductor.org/packages/release/bioc/html/proDA.html), [scp](https://www.bioconductor.org/packages/release/bioc/html/scp.html), [DEP](https://www.bioconductor.org/packages/release/bioc/html/DEP.html), and many others. They tend to share (i) statistical methodologies / algorithms (about normalisation, imputation, differential analysis and so forth) and (ii) Bioconductor formats (e.g., `QFeatures`) or low levels routines (e.g., [MsCoreUtils](https://www.bioconductor.org/packages/release/bioc/html/MsCoreUtils.html)); however, the redundancy makes it possible to find a diversity of workflows that match the specificities of each experimental design. In this landscape, `DaparToolshed` is an update of the decade-old package `DAPAR`, from the [DAPAR](https://www.bioconductor.org/packages/release/bioc/html/DAPAR.html)/[Prostar](https://www.bioconductor.org/packages/release/bioc/html/Prostar.html) suite (@Wieczorek2017-nn), which makes it possible to leverage the extended capabilities of the `QFeatures` format. However, as `DAPAR`  leverages data structure from [MSnbase](https://www.bioconductor.org/packages/release/bioc/html/MSnbase.html) that are still used nowdays in spite of `QFeatures` advent, we propose separated packages for sakes of simpler use and simpler maintenance.

Starting from a table of abundance values and associated metdata, `DaparToolshed` proposes functions to build a complete statistical analysis workflow up to the selection of differentially expressed proteins in the contrasted conditions. `DaparToolshed` makes it possible to import data at different levels: precursor, peptide or protein, each requiring specific processing.

This vignette describes a complete precursor- or peptide-level workflow.

This package is used by the `Prostar2` package, which provides a user-friendly graphical interfaces, so that programming skill is not required to achieve the data analysis.

## Installation

To install `DaparToolshed` :

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("DaparToolshed", quietly = TRUE)) pak::pak("edyp-lab/DaparToolshed")
if (!requireNamespace("Pirat", quietly = TRUE)) pak::pak("Pirat")
if (!requireNamespace("scpdata", quietly = TRUE)) pak::pak("scpdata")
```

```{r setup, message = FALSE}
library(DaparToolshed)
library(scpdata)
```


# Import dataset

For testing purposes, the `DaparToolshedData` package is accompanied with several example datasets.

The dataset used in this vignette is an extract from the `Exp1_R25_pept` dataset in the `DaparToolshedData` package (@giai_gianetto_calibration_2016). It comprises 500 peptides and 6 samples, divided into 2 conditions ("25fmol" and "10fmol"). 

```{r dataset}
dta <- dou2019_mouse()
## PREPARE DATA -----
#keep only wanted conditions
dtacond <- c("C10", "SVEC", "RAW")
colselect <- which(SummarizedExperiment::colData(dta)$SampleType %in% dtacond)

#create info for coldata
nomcoldta <- colnames(SummarizedExperiment::assay(dta[[13]][, colselect]))
nomcond <- SummarizedExperiment::colData(dta)[which(SummarizedExperiment::colData(dta)$SampleType %in% dtacond), "SampleType"]
#met tt ensemble
sampledta <- data.frame(Sample.name = nomcoldta, 
                        Condition = nomcond, 
                        Bio.Rep = seq_along(nomcoldta))
sampledta$Sample.name <- gsub(".", "_", sampledta$Sample.name, fixed = TRUE)
sampledta <- sampledta[order(sampledta$Condition), ]


#select assay
assaydta <- SummarizedExperiment::assay(dta[[13]][, colselect])
colnames(assaydta) <- gsub(".", "_", colnames(assaydta), fixed = TRUE)
rownames(assaydta) <- seq_len(nrow(assaydta))

#get rowdata
rowdta <- SummarizedExperiment::rowData(dta[[13]])
rownames(rowdta) <- seq_len(nrow(rowdta))

#create columns for metacell
metadta <- as.data.frame(replicate(ncol(assaydta), 
                                   rep("By MS/MS", nrow(assaydta)), 
                                   simplify = FALSE))
metadta <- setNames(metadta, paste0("Id_", colnames(assaydta)))

#merge together
quantdta <- cbind(assaydta, rowdta, metadta)
quantdta <- as.data.frame(quantdta)
colnames(quantdta) <- gsub(".", "_", colnames(quantdta), fixed = TRUE)

```

`DaparToolshed` uses `MultiAssayExperiment`, `SummarizedExperiment` and `QFeatures` data structures.

The function `createQFeatures()` allows for the importation of a given dataset to a `QFeatures` format. The dataset can come from any tabulated-like file containing quantitative data and metadata, accompanied by another dataframe for the sample description. The `QFeatures` format allows you to keep track of the various transformations made at each step.

For each peptide (or precursor) in each sample of the dataset, it is possible to tag the evidence that links the abundance to the molecule identified: for instance, if the abundance value is missing or imputed, if the quantification-identification match result from a transfer between samples or if there has been a fragmentation spectrum identified in relation to the quantified elution profile, etc. If such metadata is available, then the corresponding column indexes can be indicated in `indexForMetacell`. This allows the information to be leveraged during the workflow. 
The quantitative data do not need to be logged beforehand, as setting `logData = TRUE` will yield automatic log-transform. Similarly, with `force.na = TRUE`, any quantitative data with a null value or a NaN will be converted to NA. It is necessary to specify the column of the table where the mother proteins are indicated (`'parentProtId'` parameter) as an adjacency matrix is created when the function is executed. This matrix provides easy access to peptide-protein relationships, including shared peptides, i.e., when the same peptide belongs to several proteins. To run a peptidomics analysis (i.e., peptides are analysed irrespective of their mother proteins), as no peptide-to-protein aggregation is necessary, it is more suited to rely on the same workflow as for protein-level analysis.


```{r importdata, message = FALSE}
obj <- createQFeatures(data = quantdta,
                       file = 'dou2019_mouse',
                       sample = sampledta,
                       indQData = 1:72,
                       keyId = "sequence",
                       indexForMetacell = 87:158,
                       logData = TRUE,
                       force.na = TRUE,
                       typeDataset = "peptide",
                       parentProtId = "DatabaseAccess",
                       analysis = "Pept_Data",
                       software = "maxquant")

obj <- obj[1:1000,]

obj
```

```{r}
design_qf(obj)
```

The data obtained includes two assays: `original` and `logAssay`. The original, non-logged data is contained in `original`, while the logged data is in `logAssay`. This shows how the `QFeatures` format allows for recording the state of the dataset after each step. 


# Filtering

The first step in this workflow is peptide filtering.  Not all peptides in the dataset are relevant for analysis. Some may be contaminants, for example, while others may have too many missing values to be taken into account.
This filtering can be done using information contained in quantitative data tags (see above) or metadata, i.e., all the variables from the original dataset that were not quantitative data and can be accessed through the `SummarizedExperiment::rowData()` function. 

```{r metadataaccess, message = FALSE}
 #quantitative data tags
head(qMetacell(obj[[length(obj)]]))

 #metadata
head(SummarizedExperiment::rowData(obj[[length(obj)]]), n = 3)
```

To apply a filter based on tags, the `FunctionFilter()` function should be used. The functions `QFeatures::filterFeature()` and `QFeatures::VariableFilter()` can be used to create filters from metadata.

```{r filters, message = FALSE}
#create filter to remove empty lines
filter_emptyline <- FunctionFilter("qMetacellWholeLine",
                                   cmd = 'delete',
                                   pattern = 'Missing MEC')
#filter lines with too many NA is every condition
filter_manyNA <- FunctionFilter('qMetacellOnConditions',
                                cmd = 'delete',
                                mode = 'AllCond',
                                pattern = c('Missing MEC', 'Missing POV'),
                                conds = SummarizedExperiment::colData(obj)$Condition,
                                percent = TRUE,
                                th = 0.8,
                                operator = '>')
#create filter to remove contaminant
filter_contaminant <- QFeatures::VariableFilter(field = "isContaminant",
                                                value = "TRUE",
                                                condition = "==",
                                                not = TRUE)
```

Doing so creates the filter but does not apply them to the data. All created filters can be applied at once using the `filterFeaturesOneSE()` function, which will perform filtering on the assay indicated by `i` and create a new assay with the filtered peptides.

```{r filtering, message = FALSE}
 #apply all filters and create new assay
obj <- filterFeaturesOneSE(object = obj,
                           i = length(obj),
                           name = "Filtered",
                           filters = list(filter_emptyline, filter_manyNA, filter_contaminant))
```

It is advised to remove proteins that no longer have an associated peptide following this filtering from the adjacency matrix in order to facilitate the subsequent aggregation step.

```{r removeprot, message = FALSE}
 #remove proteins with no associated peptides
X <- QFeatures::adjacencyMatrix(obj[[length(obj)]])
SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix <- X[, -which(Matrix::colSums(X) == 0)]

obj
```


# Normalization

The normalization step reduces biases introduced at any preliminary stage, as to make samples more comparable. Though not a full batch effect correction method (which can be required depending on the experimental design), it can limit some simple batch effects. During this step, peptide abundances are essentially rescaled between samples of the same conditions or of all conditions.

Several algorithms are implemented in `DaparToolshed` and the list of available methods can be accessed by using `normalizeMethods()` :

```{r normalizationmethods}
 #list of available methods
normalizeMethods()
```

The methods proposed acts as described below:

- `GlobalQuantileAlignment` : Aligns the quantiles of intensity distributions across all samples. (This method should be used cautiously, as it imposes a high-impact normalization. Broadly speaking, the abundances are replaced by their order statistics).
- `SumByColumns` : Normalizes each abundance value by the total abundance of its corresponding sample. Un-logged intensities are used for processing. This method assumes the total amount of biological material is equal in all the samples .
- `QuantileCentering` : Aligns the intensity distributions to a specific quantile, e.g., the median or a lower quantile, either across all samples or within each condition.
- `MeanCentering` : Aligns the intensity distributions to their means, either across all samples or within each condition. Optionally, unit variance can be enforced.
- `LOESS` : Applies a cyclic LOESS normalization, either across all samples or within each condition. The intensity values are normalized by means of a local regression model of the difference of intensities as function of the median intensity value (@cleveland_robust_1979) (see @gentleman_limma:_2005 for implementation details).
- `vsn` : Applies the Variance Stabilization Normalization method, either across all samples or within each condition (@huber_variance_2002).


The `normalizeFunction()` function allows you to normalize data using any of the methods described above and create a new assay with normalized data. The parameters to be defined depend on the chosen method. The `type` argument allows to indicate whether the method is applied to the entire dataset at once (`"overall"`) or whether each condition is normalized independently (`"within conditions"`).

```{r normalization}
obj <- normalizeFunction(obj,
                         method = "MeanCentering",
                         scaling = TRUE,
                         type = "overall")

obj
```

The `compareNormalizationD_HC()` function provides a way to visually compare the quantitative data before and after normalization. This plot shows the influence of the normalization method.

```{r normalizationplot, message = FALSE, warning=FALSE}
obj1 <- obj[[length(obj)]]
obj2 <- obj[[length(obj)-1]]
protId <- DaparToolshed::idcol(obj1)

.n <- floor(0.02 * nrow(obj1))
.subset <- seq(nrow(obj1))

compareNormalizationD_HC(
  qDataBefore = SummarizedExperiment::assay(obj1),
  qDataAfter = SummarizedExperiment::assay(obj2),
  keyId = SummarizedExperiment::rowData(obj1)[, protId],
  conds = DaparToolshed::design_qf(obj)$Condition,
  n = .n,
  subset.view = .subset
)
```


# Imputation

It is common to face a significant number of missing values. To overcome this problem, it is often necessary to resort to imputation. However, it is important to note that data imputation amount to creating experimental measurements out of thin air (and of a bit of mathematics too). It is thus recommended to avoid imputation whenever possible. However, most algorithms required subsequently cannot deal with missing values, and the few that claim to be robust to missing values either relies on implicit or explicit imputation, which makes them no better.

First, it is interesting to look at the quantity and distribution of missing values. The `metacellPerLinesHisto_HC()` and `metacellPerLinesHistoPerCondition_HC()` functions give access to the number of rows containing a given number of missing values, respectively for all samples and within each condition. These functions use the quantitative data tags described above.

```{r missingvalueplots, warning=FALSE}
pal <- unique(GetColorsForConditions(design_qf(obj)$Condition))
pattern <- c("Missing MEC", "Missing POV")
grp <- design_qf(obj)$Condition

 #number of line with different amount of NA
metacellPerLinesHisto_HC(obj[[length(obj)]], group = grp, pattern = pattern)
 #number of line with different amount of NA per condition
metacellPerLinesHistoPerCondition_HC(obj[[length(obj)]], group = grp, pattern = pattern)
```

The `hc_mvTypePlot2()` function displays density plots showing the distribution of partially observed values for each condition. The x-axis corresponds to the mean intensity of a peptide in a condition, while the y-axis indicates the number of observed values for that peptide under the same condition.

It can be observed that the distribution tends to shift towards higher values when fewer missing values are present. This is a sign that at least some of the missing values originate from peptides that are below the lower limit of the mass spectrometer.

```{r missingvaluedistribplots}
hc_mvTypePlot2(obj[[length(obj)]], group = grp, pattern = pattern, pal = pal)
```

Numerous imputation methods exist. Here, data imputation is performed using the `wrapperPirat()` function, which is a wrapper for the `Pirat` method. `Pirat` is an imputation method that uses a penalized maximum likelihood strategy and accounts for sibling peptide correlations (@etourneau_penalized_2024). It does not require any parameter tuning. For more information on this method, see the [Pirat](https://www.bioconductor.org/packages/release/bioc/html/Pirat.html) package.

```{r imputation, message = FALSE}
obj <- wrapperPirat(data = obj,
                     adjmat = SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix,
                     extension = "base")

obj
```


# Aggregation

The next step is to aggregate the peptide-level data into protein-level data. For optimal results in the subsequent differential analysis, it is important to aggregate only after imputation. 

Beforehand, it may be useful to look at the dataset in more details and to choose the most appropriate aggregation method accordingly. The `getProteinsStats()` function displays various information about peptides and associated proteins from the adjacency matrix.

```{r aggregationadjmatrix}
X <- SummarizedExperiment::rowData(obj[[length(obj)]])$adjacencyMatrix
info_pept <- getProteinsStats(X)

message("Total number of peptides : ", info_pept$nbPeptides, "\n",
    "Number of specific peptides : ", info_pept$nbSpecificPeptides, "\n",
    "Number of shared peptides : ", info_pept$nbSharedPeptides, "\n")

message("Total number of proteins : ", info_pept$nbProt, "\n",
    "Number of protein with only specific peptides : ", length(info_pept$protOnlyUniquePep), "\n",
    "Number of protein with only shared peptides : ", length(info_pept$protOnlySharedPep), "\n",
    "Number of protein with both : ", length(info_pept$protMixPep), "\n")
```

There are predominantly specific peptides, but several proteins contain shared peptides. The question of how to manage these shared peptides may arise. Contrarily to most aggregation functions, `RunAggregation()` allows shared peptides to be taken into account in various ways --to be specified using the `includeSharedPeptides` argument. First, they can be handled as specific peptides (`Yes_As_Specific`), in which case they will be duplicated for each protein they belong to. Second, they can have their intensity redistributed proportionally across the different proteins (`Yes_Iterative_Redistribution` or `Yes_Simple_Redistribution`). Finally, shared peptides can be ignored (`No`). If `Yes_Iterative_Redistribution`, `Yes_Simple_Redistribution` or `No` is chosen, proteins containing only shared peptides will have no associated value. 
In addition, there is the choice of aggregation function, used for quantitative feature aggregation, with `operator`. This can be either `Sum`, `Mean`, `Median`, `medianPolish` or `robustSummary`.

```{r aggregation, message = FALSE}
obj <- RunAggregation(obj,
                      adjMatrix = 'adjacencyMatrix',
                      includeSharedPeptides = 'Yes_Iterative_Redistribution',
                      operator = 'Mean',
                      considerPeptides = 'allPeptides',
                      ponderation = "Global",
                      max_iter = 500
                      )

obj
```

After aggregation, it can be seen that there are still some missing values. These come from proteins that only have shared peptides, which have been left out using this method, as explained above. 

```{r aggregationsummary}
summary(SummarizedExperiment::assay(obj[["aggregated"]]))
```

A new filter can thus be applied to these protein-level data to remove them.

```{r filteringprot}
 #apply the filter to remove empty lines created
obj <- filterFeaturesOneSE(object = obj,
                           i = length(obj),
                           name = "FilterProt",
                           filters = list(filter_emptyline))

obj
```


# Differential Analysis

Differential protein expression analysis compares the relative abundance of the same protein between two conditions. The comparison between two conditions results in what is called a contrast, for example “A vs. B.” 
The dataset used here contains only two conditions, hence there is only one contrast to be considered. 

For additional advice on this step, see @wieczorek_five_2019.

## Hypothesis test : 2 conditions to compare and selection on Volcano plot


Cette section se place dans le cas d'une comparaison de deux groupes uniquement (sans autres groupes ni covariable). 
Dans l'analyse différentielle, il peut être pertinent de faire une sélection sur un Volcano Plot afin de garder les protéines ayant de petites $p$-valeurs et de grands foldChange. 

Le controle du FDR (espérance de la Proportion de Faux Positif (FDP)) n'est pas garantie dans le cas d'une telle sélection (voir Ebrahimpour et al ...[???]). 
Une alternative est de contrôler la médiane de la FDP. Pour cela nous allons utiliser la méthode **sanssouci** qui donne des bornes dites post hoc sur la FDP dans *des sous-ensembles d'hypothèses arbitraires, multiples et éventuellement guidés par des données*. 

Pour cela nous allons utiliser le package `sanssouci` qui permet de (1) calculer les $p$-valeurs et le logFoldChange selon le test choisi, (2) calculer les guaranties post hoc sur le FDP et (3) obtenir les bornes post hoc sur une sélection faite à postériori sur le volcano plot (associées aux p-valeurs et foldchange de l'étape (1)). 

Dans le cas de l'étude différentielle de protéine, le package `Limma` est utilisé. 

### Initialisation of SansSouci object

La première étape est d'initialiser l'objet `SansSouci`. Attention, ici les groupes doivent être codé avec des labels '1' et '0' pour les deux labels de groupes à comparer.

```{r}
library(sanssouci4DaparToolshed)
library(sanssouci)

# selectionne 2 conditions
indexcond <- which(colData(obj)$Condition %in% c("C10", "RAW"))
obj_selectcond <- obj[, indexcond]

sanssouci_object <- wrapper_QFtoSansSouci(obj_selectcond)
```

### Hypothesis testing and post hoc calibration 

[ici je n'aime pas le titre parce que je parle de calibration mais ce n'est jamais introduit avant]

Il faut définir les paramètres du test. 
Pour contrôler la médianne de la FDP il faut choisir le niveau $\alpha = 0.5$.

[Ici on pourrait discuter de la valeur de B? Par défaut elle vaut 1000 dans notre code.]

```{r}
alpha = 0.5
```


Le calcul des pvaleurs ainsi que la calibration des bornes post hoc se fait avec la fonction `fit` du package `sanssouci` en fournissant l'objet SansSouci, le niveau d'erreur $\alpha$ ainsi que la fonction de test. 

```{r}
sanssouci_object_cal_limma <- fit(sanssouci_object, alpha = alpha, 
                                  rowTestFUN = rowLimmaTest, B = 1000)
sanssouci_object_cal_limma
```

A partir de cet objet il est possible d'extraire les pvaleurs et le logFolchange par les fonction d'appel `pValues` et `foldChanges` respectivement. 

```{r}
pval <- as.vector(pValues(sanssouci_object_cal_limma))
logFC <- as.vector(foldChanges(sanssouci_object_cal_limma))
```


### Volcano plot and post hoc bound

Il est possible d'avoir une garantie post hoc sur n'importe quelle sélection faite sur le vocanoplot en appliquant un seuil sur les pvaleurs ainsi que le foldchange. 

```{r}
pval_threshold <- 0.01
logfc_threshold <- 0.04

volcanoPlot(x = sanssouci_object_cal_limma,  
                   pval_thr = pval_threshold, 
                   logfc_thr = logfc_threshold, interactive = FALSE)
```

La filtration permets de sélectionner 39 protéines. Parmis ces protéines, la garanties statistiques à 50% permet de dire qu'il y a au moins 35 protéines qui sont réellement différentiellement exprimées. Cela revient à dire que la proportion de faux positifs n'excède pas 10%.

[heu on a forcé le volcano plot a utiliser le terme gène. On refait avec ggplot2 ?]

### Extraction des protéines sélectionnées 

Proposition d'extraction et sorti : tableau data.frame qui contient toutes les protéines, les pvaleurs associées ainsi que les logFC. Il y a autant de colonnes en suivant que de sélection. Une colonne de sélection est consititué de 1 pour la sélection et 0 pour la non sélection. Cette colonne a des attributs reprenant les seuils de sélection + les bornes post hoc FDP et TP. 


```{r}
obj_selectcond <- export2dapartoolshed(sanssouci_obj = sanssouci_object_cal_limma, 
                            pval_thr = pval_threshold, 
                            logfc_thr = logfc_threshold, 
                            qfeature_obj = obj_selectcond)

head(metadata(obj_selectcond)$posthoc_selection)
```

Pour obtenir les bornes post hoc 

```{r}
getPostHocBound(obj_selectcond, "sel_1")
```


# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
